Meta-learning in movement prediction problem of aperiodic time-series data

### 2024-12-08 doc/paper/paper.tex
paper vn

\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{rotating}
\usepackage{multirow}

@@

    \textbf{Abstract}. Predicting aperiodic time-series data (e.g. stock price, foreign exchange, Bitcoin price,...) is a difficult task for machine learning models because this type of data has high and not stationary variance; does not show clearly periodicity, making it difficult to extract features; depends not only on past values but also on external factors, which make them unstable and non-cyclical, such as economic and political situations. To overcome the above challenges, we use Meta-learning to train a combined \verb|LSTM| and \verb|CNN| network, thereby effectively extracting and synthesizing hidden features of the data over time. Experiments on foreign exchange data of 60 currency pairs over 24 years (2000-2024) show that the proposed method performs well and has 5\%-11\% higher accuracy compared to the \verb|NHITS| - the state-of-the-art model in 2023 on time-series data, in the task of predicting the trend (upward or downward) of the next trading day.

    % Dự đoán aperiodic time-series data (e.g. stock price, foreign exchange, Bitcoin price,...) là một tác vụ khó khăn đối với các mô hình học máy bởi vì loại dữ liệu này có phương sai cao và không cố định; không thể hiện chu kỳ rõ ràng nên khó rút trích đặc trưng; không chỉ phụ thuộc vào giá trị quá khứ mà còn phụ thuộc các yếu tố ngoại lai, điều khiến chúng bất ổn và không có chu kỳ như tình hình kinh tế, chính trị. Để khắc phục các thách thức nêu trên, chúng tôi sử dụng Meta-learning để huấn luyện kết hợp mạng \verb|LSTM| và \verb|CNN|, từ đó rút trích và tổng hợp hiệu quả các đặc trưng ẩn của dữ liệu theo thời gian. Thực nghiệm trên dữ liệu foreign exchange của 15 cặp tiền tệ trong vòng 24 năm (2000-2024) cho thấy phương pháp đề xuất hoạt động tốt và có độ chính xác cao hơn mô hình \verb|NHITS| - state-of-the-art model của năm 2023 trên dữ liệu time-series, trong bài toán dự đoán xu hướng (upward or downward) của ngày giao dịch tiếp theo.
\end{abstract}
%----------------------------------------
\keywords{Meta-learning, \verb|LSTM|, \verb|CNN|, Aperiodic time-series data, Foreign exchange}
\maketitle

@@

% Aperiodic time-series prediction nói chung hay foreign exchange (FX) prediction nói riêng từ lâu đã là vấn đề đáng quan tâm của nhiều nghiên cứu \citep{li2019multi, islam2021foreign, heryadi2021foreign}. Hai kỹ thuật chính được sử dụng trong aperiodic time-series prediction là fundamental analysis and technical analysis \cite{ayitey2023forex}. Trong khi fundamental analysis thiên về phân tích các yếu tố tác động từ bên ngoài và khó có thể capture từ các biến thiên giá trị trong quá khứ như chính sách, chiến lược kinh tế của công ty, quốc gia để dự đoán tương lai; technical analysis dựa hoàn toàn vào lịch sử biến động giá trị để phân tích xu hướng tương lai.

Aperiodic time-series prediction in general or foreign exchange (FX) prediction in particular has been a matter of concern for many researchers \citep{li2019multi, islam2021foreign, heryadi2021foreign} for a long time. The two main techniques used in aperiodic time-series prediction are fundamental analysis and technical analysis \cite{ayitey2023forex}. Fundamental analysis focuses on analyzing external factors such as policies and economic strategies of companies and countries to predict the future. Meanwhile, technical analysis relies entirely on historical value fluctuations, which are difficult to capture external factors, to analyze future trends.

@@

% Việc dự đoán trên aperiodic time-series data gặp một vài thách thức cố hữu. Trong đó có thể kể tới: (1) - Variance của loại dữ liệu này biến thiên mạnh qua các thời kỳ. Từ đó giả thuyết rằng chúng tuân theo một phân phối để có thể xấp xỉ lỗi là không thể sử dụng được, dẫn đến việc các mô hình học máy khó có thể dự đoán chính xác giá trị cũng như xu hướng trong tương lai; (2) - Aperiodic data không tuân theo bất kỳ quy luật rõ ràng nào nên việc học các đặc trưng ẩn của dữ liệu để tiến hành dự đoán gặp rất nhiều khó khăn; (3) - Aperiodic time-series data (e.g. giá cổ phiếu của một công ty) không hoàn toàn phụ thuộc vào dữ liệu quá khứ mà còn phụ thuộc nhiều yếu tố ngoại lai (e.g. tin tức, tình hình kinh tế, chính trị) \cite{li2019multi}.

Aperiodic time-series data prediction faces several inherent challenges: (1) - The variance varies significantly over time, making it difficult to assume a specific distribution to approximate errors, which is a challenge for machine learning models in accurately predicting future values and trends. (2) - Aperiodic data does not follow explicit patterns, making it challenging to identify the latent features required for reliable predictions. (3) - Aperiodic time-series data is influenced not only by historical data but also by numerous external factors \cite{li2019multi}. For example, a company's stock price is affected by news, economic conditions, political events, and other external influences.

@@

% Đối với thách thức đầu tiên, các mô hình ensemble [cite something here] thường được sử dụng để hạn chế ảnh hưởng của sự biến đổi variance. Ensemble model giúp cung cấp một cái nhìn tổng quát, tổng hợp từ nhiều khía cạnh dựa trên các sub-models, từ đó giúp mô hình tổng quát thích ứng được với sự thay đổi mạnh của variance.
% Tuy nhiên, dưới sự tổng hợp cứng nhắc của các mô hình ensemble truyền thống, kết quả thu được trên các bộ dữ liệu phi chi kỳ là rất thấp.

For the variance variation challenge, ensemble models \citep{sadeghi2021combined, zafeiriou2020intraday, ali2020complete} are often used to mitigate the effects of variance variation. Ensemble models provide a holistic and multi-perspective view based on sub-models, thereby helping the overall model adapt to strong variance changes. However, under the rigid synthesis of traditional ensemble models, the results obtained on non-periodic datasets are very poor.

@@

% Để khắc phục thách thức thứ hai, các nghiên cứu phần lớn sử dụng các đặc trưng rút trích được từ Long short-term memory neural network (\verb|LSTM|) \cite{hochreiter1997long}, Artificial neural network (\verb|ANN|), và Convolution neural network (\verb|CNN|) \cite{lecun1989handwritten}. Cụ thể, trong năm 2022, 20\% tổng số các bài báo liên quan đến dự đoán chỉ số tài chính sử dụng \verb|LSTM|, 20\% sử dụng \verb|ANN| và 6\% sử dụng \verb|CNN| \cite{ayitey2023forex}.
% Điều này là hoàn toàn dễ hiểu vì \verb|LSTM| và \verb|CNN| đã chứng minh được khả năng của mình trong việc rút trích các đặc trưng không gian, thời gian một cách hiệu quả.
% Mặc dù vậy, khi đối mặt với tính phi chu kỳ, sau khoảng 100 training steps, \verb|LSTM| vẫn bị underfitting, còn \verb|CNN| bị overfitting chỉ sau 5 training steps

To overcome the pattern variation challenge, most studies use features extracted from Long Short-Term Memory neural network (\verb|LSTM|) \cite{hochreiter1997long}, Artificial Neural Network (\verb|ANN|), and Convolution Neural Network (\verb|CNN|) \cite{lecun1989handwritten}. Specifically, in 2022, 20\% of all publications related to financial index prediction used \verb|LSTM|, 20\% used \verb|ANN|, and 6\% used \verb|CNN| \cite{ayitey2023forex}. This is completely understandable because \verb|LSTM| and \verb|CNN| have proven their ability to extract spatial and temporal features effectively. However, when faced with aperiodic data with only 2600 samples, after 40 training steps, \verb|LSTM| is still underfitting, while \verb|CNN| is overfitting after only 5 training steps.

@@

% Đối với thách thức thứ ba, nghiên cứu \cite{fama1970efficient} đưa ra giả thuyết rằng các time-series datasets khác nhau của cùng một lĩnh vực tại cùng thời điểm phản ánh sự tác động của các yếu tố ngoại lai. Ví dụ, các nghiên cứu \citep{overreactioncontrarian, mech1993portfolio} cũng chỉ ra sự phụ thuộc giữa chỉ số tài chính của một công ty nhất định và các chỉ số của các công ty khác. Điều này càng làm tăng tính đúng đắn của giả thuyết trong nghiên cứu \cite{fama1970efficient}.
% Do đó, bằng việc khai thác các nguồn dữ liệu có cùng domain, chúng ta có thể capture được các thông tin phản ánh các external factors.

Regarding the external influences challenge, \cite{fama1970efficient} hypothesizes that different time-series datasets of the same domain at the same time reflect the impact of external factors. For example, studies \citep{overreactioncontrarian, mech1993portfolio} show the dependence between the financial ratios of a given company and the ratios of other companies. This further strengthens the hypothesis in study \cite{fama1970efficient}. Therefore, by exploiting data sources with the same domain, we can capture information reflecting external factors.

@@

% Chúng tôi tiếp cận các thách thức nêu trên thông qua việc kết hợp đặc trưng và tổng hợp mô hình. Cụ thể, bằng cách sử dụng Meta-learning (ML) \cite{finn2017model}, chúng tôi tổng hợp hiệu quả tham số của các mô hình cục bộ, giúp giảm thiểu đáng kể variance loss cũng như tăng khả năng thu thập các thông tin phản ánh external factors từ các tập dữ liệu có cùng domain. Để học được các laten patterns, chúng tôi đề xuất phương pháp kết hợp các đặc trưng từ \verb|CNN| và \verb|LSTM|. Ngoài ra, chúng tôi cho rằng, aperiodic time-series data còn có những phụ thuộc ngầm vào các thời điểm nhất định trong quá khứ (hidden long-term dependency). Đối với cách tiếp cận truyền thống, người ta sử dụng một lượng dữ liệu quá khứ cố định (lookback window) để huấn luyện mô hình. Điều này gây một trở ngại lớn cho quá trình học vì các đặc trưng dài hạn theo thời gian sẽ bị quên. Mặt khác, ML chia nhỏ tập dữ liệu thành nhiều phần để học và tổng hợp hiệu quả các tham số học được nên có thể xử lý tốt thách thức này.

We approach the above challenges through combination of feature combination and model synthesis. Specifically, by using Meta-learning (ML) \cite{finn2017model}, we efficiently synthesize the parameters of local models, which significantly reduces variance loss and increases the ability to collect information reflecting external factors from datasets in the same domain. To effectively learn laten patterns, we propose a method that combines features from \verb|CNN| and \verb|LSTM|. In addition, we assume that aperiodic time-series data also has hidden long-term dependencies. Traditional approaches use a fixed amount of past data (lookback window) to train the model. This poses a major obstacle to the learning process because long-term features will be forgotten over time. On the other hand, ML divides the dataset into many parts to learn and synthesize the learned parameters effectively, so it can handle this challenge well.

@@

% Cuối cùng, chúng tôi chứng minh tính ưu việt của thuật toán đề xuất bằng cách giải bài toán dự đoán xu hướng (tăng hoặc giảm) của tỉ giá ngoại hối và so sánh kết quả với mô hình SOTA hiện tại (\verb|NHITS| \cite{challu2023nhits}) trên hai loại dữ liệu: (1) - Dữ liệu tỉ giả của cặp tiền tệ USD/JPY; (2) - Dữ liệu tỷ giá của 60 cặp tiền tệ, cấu thành từ 18 quốc gia. Các tập dữ liệu này được công khai trên Internet và có thể tải về dễ dàng. Bản cài đặt chính thức có thể xem tại [bỏ cái link github vào đây].

Finally, we demonstrate the superiority of the proposed algorithm by solving the problem of predicting the trend (upward or downward) of foreign exchange rates and comparing the results with the existing state-of-the-art (SOTA) model (\verb|NHITS| \cite{challu2023nhits}) on two types of data: (1) - USD/JPY exchange rate data; (2) - Exchange rate data of 60 currency pairs, comprising 18 countries. These data sets are publicly available on the Internet and can be easily downloaded. The official implementation can be found at \url{https://github.com/baolongnguyenmac/multi_fx}.

@@

In summary, our main contributions include the following: First, we present a novel \textbf{feature combination technique} for aperiodic time-series data that extracts and combines features using both \verb|LSTM| and \verb|CNN| models. Second, we uncover hidden long-term dependencies, demonstrating experimentally that aperiodic time-series data not only rely on external factors but also exhibit dependencies with past period, which we call \textbf{hidden long-term dependencies}. Third, we introduce an \textbf{efficient model parameter aggregation approach} by employing ML algorithm in place of traditional ensemble models for combining parameters from multiple machine learning models. Finally, we validate our method through experiments on exchange rate datasets and show its effectiveness in comparison to the SOTA model, \verb|NHITS|.

@@

% Như đã đề cập, \verb|LSTM| là mạng neural rất phổ biến trong việc handle các bài toán liên quan đến dự đoán trên dữ liệu time-series. \verb|LSTM| được sử dụng phổ biến như vậy bởi vì nó xử lý tốt vấn đề vanishing gradient (dễ dàng bắt gặp khi sử dụng Recurrent neural network) và có thể khai thác hiệu quả các mối quan hệ phi tuyến trong dữ liệu. Thật vậy, bằng cách duy trì cell-state trong mỗi iteration, \verb|LSTM| có thể khắc phục vấn đề vanishing gradient, từ đó bảo toàn khả năng capture các phụ thuộc dài hạn \cite{cheng2018applied}. Ngoài ra, \verb|LSTM| thực hiện rút trích đặc trưng với các hàm kích hoạt phi tuyến, giúp các tham số mô hình có thể capture tính phi tuyến của dữ liệu \cite{he2016exploiting}. Hai yếu tố nêu trên khiến cho \verb|LSTM| trở thành lựa chọn đầu tiên được nghĩ đến khi giải quyết các bài toán trên dữ liệu time-series.

As aforementioned, \verb|LSTM| is a well-known neural network for handling prediction problems on time-series data. \verb|LSTM| is commonly used because it handles the problem of vanishing gradients well (easily encountered when using Recurrent neural networks) and can effectively exploit nonlinear relationships in data. Indeed, by maintaining the cell-state in each iteration, \verb|LSTM| can overcome the vanishing gradient problem, thereby preserving the ability to capture long-term dependencies \cite{cheng2018applied}. In addition, \verb|LSTM| performs feature extraction with nonlinear activation functions, which helps the model parameters capture the nonlinearity of the data \cite{he2016exploiting}. These factors make \verb|LSTM| to be the first choice to think of when solving problems on time-series data.

@@

% Mạng \verb|CNN| được sử dụng rất nhiều trong các tác vụ xử lý hình ảnh \citep{naranjo2020review, sharma2018analysis} bởi khả năng tổng hợp các quan hệ cục bộ. Không chỉ vậy, \verb|CNN| còn được dùng rất nhiều trong các tác vụ xử lý dữ liệu time-series như speech recognition \cite{dua2022developing}, natural language processing \cite{varshitha2023natural}. Điều đó chứng tỏ được khả năng của \verb|CNN| trong việc khám phá mối quan hệ thời gian cục bộ giữa các mẫu dữ liệu. Mặc dù vậy, \verb|CNN| lại rất ít được dùng trong các tác vụ dự đoán stock price hay foreign exchange. Trong nghiên cứu này, chúng tôi tận dụng khả năng rút trích đặc trưng cục bộ tuyệt vời của \verb|CNN| để tích hợp thêm thông tin ẩn vào quá trình huấn luyện của mô hình.

\verb|CNN| is widely used in image processing tasks \citep{naranjo2020review, sharma2018analysis} because of its ability to synthesize local relationships. Not only that, \verb|CNN| is also widely used in time-series data processing tasks such as speech recognition \cite{dua2022developing}, natural language processing \cite{varshitha2023natural}. This proves the ability of \verb|CNN| in discovering local temporal relationships between data samples. However, \verb|CNN| is rarely used in aperiodic time-series data prediction tasks. In this study, we take advantage of \verb|CNN|'s excellent local feature extraction ability to incorporate more hidden information into the model training process.

@@

% Các thuật toán Meta-learning (ML), điển hình là MAML \cite{finn2017model} được biết đến với khả năng huấn luyện một mô hình có tính tổng quát cao, thích ứng nhanh trên tập dữ liệu mới thông qua một lượng nhỏ dữ liệu và số bước huấn luyện \citep{hospedales2021meta, vettoruzzo2024advances}. Với khả năng này, ML được sử dụng rất nhiều trong các tác vụ đòi hỏi khả năng đáp ứng của mô hình trên dữ liệu (e.g. cá nhân hóa mô hình học \citep{chen2018federated, fallah2020personalized,nguyen2022meta}, domain adaptation trong online learning \citep{hu2023meta, khoee2024domain}).

>>>>

\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{tabularray}
% \usepackage{color}

@@

    % \textbf{Abstract}. Predicting aperiodic time-series data (e.g. stock price, foreign exchange, Bitcoin price, etc.) is a difficult task for machine learning models because this type of data has high and not stationary variance; does not show clearly periodicity, making it difficult to extract features; depends not only on past values but also on external factors such as economic and political situations, which make them unstable and non-cyclical. To overcome the above challenges, we use Meta-learning to train a combined \verb|LSTM| and \verb|CNN| network, thereby effectively extracting and synthesizing deep hidden features of the data over time. Experiments on aperiodic data (foreign exchange data of 60 currency pairs over 24 years from 2000 to 2024) show that the proposed method performs well and has 5\%-11\% higher accuracy compared to the \verb|NHITS| - the state-of-the-art model in 2023 on time-series data, in the task of predicting the trend (upward or downward) of the next trading day. Meanwhile, the results obtained on periodic data (Electricity Transformer Temperature, Weather) are approximately \verb|NHITS|.

    \textbf{Abstract}. Dự đoán dữ liệu chuỗi thời gian phi chu kỳ (ví dụ: giá cổ phiếu, tỷ giá ngoại hối, giá Bitcoin,...) là một tác vụ khó khăn đối với các mô hình học máy bởi vì loại dữ liệu này có phương sai cao và không cố định; không thể hiện chu kỳ rõ ràng nên khó rút trích đặc trưng; không chỉ phụ thuộc vào giá trị quá khứ mà còn phụ thuộc các tác động bên ngoài như tình hình kinh tế, chính trị, khiến chúng bất ổn và không có chu kỳ. Để khắc phục các thách thức nêu trên, chúng tôi sử dụng Meta-learning để huấn luyện kết hợp mạng \verb|LSTM| và \verb|CNN|, từ đó rút trích và tổng hợp hiệu quả các đặc trưng ẩn của dữ liệu theo thời gian. Các thử nghiệm trên dữ liệu phi chu kỳ (dữ liệu ngoại hối của 60 cặp tiền tệ trong 24 năm từ 2000 đến 2024) cho thấy phương pháp đề xuất hoạt động tốt và có độ chính xác cao hơn 5\%-11\% so với \verb|NHITS| - mô hình SOTA 2023 trên dữ liệu chuỗi thời gian, trong nhiệm vụ dự đoán xu hướng (tăng hoặc giảm) của ngày giao dịch tiếp theo. Trong khi đó, kết quả thu được trên dữ liệu chu kỳ (Nhiệt độ máy biến áp điện, Thời tiết) đạt xấp xỉ \verb|NHITS|.
\end{abstract}
%----------------------------------------
\keywords{Meta-learning, \verb|LSTM|, \verb|CNN|, Aperiodic time-series data, Foreign exchange}
\maketitle

@@

Dự đoán dữ liệu chuỗi thời gian phi chu kỳ nói chung hay dự đoán tỷ giá ngoại hối (foreign exchange, FX) nói riêng từ lâu đã là vấn đề đáng quan tâm của nhiều nghiên cứu \citep{li2019multi, islam2021foreign, heryadi2021foreign}. Hai kỹ thuật chính được sử dụng trong aperiodic time-series prediction là fundamental analysis and technical analysis \cite{ayitey2023forex}. Trong khi fundamental analysis thiên về phân tích các yếu tố tác động từ bên ngoài và khó có thể capture từ các biến thiên giá trị trong quá khứ như chính sách, chiến lược kinh tế của công ty, quốc gia để dự đoán tương lai; technical analysis dựa hoàn toàn vào lịch sử biến động giá trị để phân tích xu hướng tương lai.

% Aperiodic time-series prediction in general or foreign exchange (FX) prediction in particular has been a matter of concern for many researchers \citep{li2019multi, islam2021foreign, heryadi2021foreign} for a long time. The two main techniques used in aperiodic time-series prediction are fundamental analysis and technical analysis \cite{ayitey2023forex}. Fundamental analysis focuses on analyzing external factors such as policies and economic strategies of companies and countries to predict the future. Meanwhile, technical analysis relies entirely on historical value fluctuations, which are difficult to capture external factors, to analyze future trends.

@@

Theo \cite{li2019multi}, việc dự đoán trên aperiodic time-series data gặp một vài thách thức cố hữu. Trong đó có thể kể tới: (1) - Variance của loại dữ liệu này biến thiên mạnh qua các thời kỳ. Từ đó giả thuyết rằng chúng tuân theo một phân phối để có thể xấp xỉ lỗi là không thể sử dụng được, dẫn đến việc các mô hình học máy khó có thể dự đoán chính xác giá trị cũng như xu hướng trong tương lai; (2) - Aperiodic data không tuân theo bất kỳ quy luật rõ ràng nào nên việc học các đặc trưng ẩn của dữ liệu để tiến hành dự đoán gặp rất nhiều khó khăn; (3) - Aperiodic time-series data (e.g. giá cổ phiếu của một công ty) không hoàn toàn phụ thuộc vào dữ liệu quá khứ mà còn phụ thuộc nhiều yếu tố ngoại lai (e.g. tin tức, tình hình kinh tế, chính trị).

% Aperiodic time-series data prediction faces several inherent challenges: (1) - The variance varies significantly over time, making it difficult to assume a specific distribution to approximate errors, which is a challenge for machine learning models in accurately predicting future values and trends. (2) - Aperiodic data does not follow explicit patterns, making it challenging to identify the latent features required for reliable predictions. (3) - Aperiodic time-series data is influenced not only by historical data but also by numerous external factors \cite{li2019multi}. For example, a company's stock price is affected by news, economic conditions, political events, and other external influences.

@@

Đối với thách thức đầu tiên, các mô hình ensemble \citep{sadeghi2021combined, zafeiriou2020intraday, ali2020complete} thường được sử dụng để hạn chế ảnh hưởng của sự biến đổi variance. Ensemble model giúp cung cấp một cái nhìn tổng quát, tổng hợp từ nhiều khía cạnh dựa trên các sub-models, từ đó giúp mô hình tổng quát thích ứng được với sự thay đổi mạnh của variance. Tuy nhiên, dưới sự tổng hợp cứng nhắc của các mô hình ensemble truyền thống, kết quả thu được trên các bộ dữ liệu phi chu kỳ là rất thấp.

% Ensemble models \citep{sadeghi2021combined, zafeiriou2020intraday, ali2020complete} are often used to mitigate the effects of variance variation challenge. Ensemble models provide a holistic and multi-perspective view based on sub-models, thereby helping the overall model adapt to strong variance changes. However, under the rigid synthesis of traditional ensemble models, the results obtained on non-periodic datasets are very poor.

@@

Để khắc phục thách thức thứ hai, các nghiên cứu phần lớn sử dụng các đặc trưng rút trích được từ Long short-term memory neural network (\verb|LSTM|) \cite{hochreiter1997long}, Artificial neural network (\verb|ANN|), và Convolution neural network (\verb|CNN|) \cite{lecun1989handwritten}. Cụ thể, trong năm 2022, 20\% tổng số các bài báo liên quan đến dự đoán chỉ số tài chính sử dụng \verb|LSTM|, 20\% sử dụng \verb|ANN| và 6\% sử dụng \verb|CNN| \cite{ayitey2023forex}. Điều này là hoàn toàn dễ hiểu vì \verb|LSTM| và \verb|CNN| đã chứng minh được khả năng của mình trong việc rút trích các đặc trưng không gian, thời gian một cách hiệu quả. Mặc dù vậy, khi đối mặt với tính phi chu kỳ trên 2600 mẫu dữ liệu FX, sau khoảng 100 training steps, \verb|LSTM| vẫn bị underfitting, còn \verb|CNN| bị overfitting chỉ sau 5 training steps

% To overcome the pattern variation challenge, most studies use features extracted from Long Short-Term Memory neural network (\verb|LSTM|) \cite{hochreiter1997long}, Artificial Neural Network (\verb|ANN|), and Convolution Neural Network (\verb|CNN|) \cite{lecun1989handwritten}. Specifically, in 2022, 20\% of all publications related to financial index prediction used \verb|LSTM|, 20\% used \verb|ANN|, and 6\% used \verb|CNN| \cite{ayitey2023forex}. This is completely understandable because \verb|LSTM| and \verb|CNN| have proven their ability to extract spatial and temporal features effectively. However, when faced with aperiodic data with only 2600 samples, after 40 training steps, our experiment shows that \verb|LSTM| is still underfitting, while \verb|CNN| is overfitting after only 5 training steps.

@@

Đối với thách thức thứ ba, nghiên cứu \cite{fama1970efficient} đưa ra giả thuyết rằng các time-series datasets khác nhau của cùng một lĩnh vực tại cùng thời điểm phản ánh sự tác động của các yếu tố ngoại lai. Các nghiên cứu \citep{overreactioncontrarian, mech1993portfolio} cũng chỉ ra sự phụ thuộc giữa chỉ số tài chính của một công ty nhất định và các chỉ số của các công ty khác. Điều này càng làm tăng tính đúng đắn của giả thuyết trong nghiên cứu \cite{fama1970efficient}. Do đó, bằng việc khai thác các nguồn dữ liệu có cùng domain, chúng ta có thể capture được các thông tin phản ánh tác động bên ngoài.

% Regarding the challenge of external influences, \cite{fama1970efficient} hypothesizes that different time-series datasets of the same domain at the same time reflect the impact of external factors. Studies \citep{overreactioncontrarian, mech1993portfolio} also show the dependence between the financial ratios of a given company and the ones of other companies. This further strengthens the hypothesis in study \cite{fama1970efficient}. Therefore, by exploiting data sources with the same domain, we can capture information reflecting external factors.

@@

Chúng tôi tiếp cận các thách thức nêu trên thông qua việc kết hợp đặc trưng và tổng hợp mô hình. Cụ thể, bằng cách sử dụng Meta-learning (ML) \cite{finn2017model}, chúng tôi tổng hợp hiệu quả tham số của các mô hình cục bộ, giúp giảm thiểu đáng kể variance loss cũng như tăng khả năng thu thập các thông tin phản ánh tác động bên ngoài từ các tập dữ liệu có cùng domain. Để học được các laten patterns, chúng tôi đề xuất phương pháp kết hợp các đặc trưng từ \verb|CNN| và \verb|LSTM|. Ngoài ra, chúng tôi cho rằng, aperiodic time-series data còn có những phụ thuộc ngầm vào các thời điểm nhất định trong quá khứ (hidden long-term dependency). Đối với cách tiếp cận truyền thống, người ta sử dụng một lượng dữ liệu quá khứ cố định (lookback window) để huấn luyện mô hình. Điều này gây một trở ngại lớn cho quá trình học vì các đặc trưng dài hạn theo thời gian sẽ bị quên. Mặt khác, ML chia nhỏ tập dữ liệu thành nhiều phần để học và tổng hợp hiệu quả các tham số học được nên có thể xử lý tốt thách thức này.

% We approach the above challenges through the combination of feature and model synthesis. Specifically, by using Meta-learning (ML) \cite{finn2017model}, we efficiently synthesize the parameters of local models, which significantly reduces variance loss and increases the ability to collect information reflecting external factors from datasets in the same domain. To effectively learn laten patterns, we propose a method that combines features from \verb|CNN| and \verb|LSTM|. In addition, we assume that aperiodic time-series data also has hidden long-term dependencies. Traditional approaches use a fixed amount of past data (lookback window) to train the model. This poses a great obstacle to the learning process because long-term features will be forgotten over time. On the other hand, ML divides the dataset into many parts to learn and synthesize the learned parameters effectively, so it can handle this challenge well.

@@

Cuối cùng, chúng tôi chứng minh tính ưu việt của thuật toán đề xuất bằng cách giải quyết bài toán dự đoán xu hướng (tăng hoặc giảm) của dữ liệu chuỗi thời gian và so sánh kết quả với mô hình SOTA năm 2023 (\verb|NHITS| \cite{challu2023nhits}) trên hai loại dữ liệu: (1) - Dữ liệu phi chu kỳ, bao gồm dữ liệu tỷ giá hối đoái USD/JPY và dữ liệu tỷ giá hối đoái của 60 cặp tiền tệ, bao gồm 18 quốc gia từ năm 2000 đến năm 2024; (2) - Dữ liệu có chu kỳ bao gồm tập dữ liệu Electricity Transformer Temperature \cite{zhou2021informer} và Weather \cite{Kolle}. Các tập dữ liệu này được cung cấp công khai trên Internet và có thể tải xuống dễ dàng. Bản cài đặt chính thức có thể xem tại \url{https://github.com/baolongnguyenmac/multi_fx}.

% Finally, we demonstrate the superiority of the proposed algorithm by solving the problem of predicting the trend (upward or downward) of time-series data and comparing the results with the existing state-of-the-art (SOTA) model (\verb|NHITS| \cite{challu2023nhits}) on two types of data: (1) - Aperiodic data, which contains USD/JPY exchange rate data and the exchange rate data of 60 currency pairs, comprising 18 countries from 2000 to 2024; (2) - Periodic data which contains Electricity Transformer Temperature \cite{zhou2021informer} and Weather \cite{Kolle} dataset. These datasets are publicly available on the Internet and can be downloaded easily. The official implementation can be found at \url{https://github.com/baolongnguyenmac/multi_fx}.

@@

Tóm lại, đóng góp chính của chúng tôi như sau:

\begin{itemize}
    \item Kết hợp đặc trưng: Rút trích đặc trưng bằng cách kết hợp các đặc trưng của \verb|LSTM| và \verb|CNN|.
    \item Hidden long-term dependency: Chứng minh thực nghiệm rằng dữ liệu time-series tồn tại các phụ thuộc ẩn với chính nó tại nhiều thời điểm khác nhau trong quá khứ.
    \item Tổng hợp hiệu quả tham số mô hình: Sử dụng ML thay cho các mô hình ensemble truyền thống trong việc tổng hợp kết quả từ các mô hình học máy.
    \item Experiment: Thực nghiệm trên dữ liệu phi chu kỳ (tỷ giá hối đoái của 60 cặp tiền tệ) và dữ liệu có chu kỳ (Electricity Transformer Temperature, Weather) rồi so sánh với mô hình SOTA 2023 \verb|NHITS| để chứng minh tính hiệu quả của phương pháp đề xuất.
\end{itemize}

% In summary, our main contributions include the following: First, we present a novel \textbf{feature combination technique} for aperiodic time-series data that extracts and combines features using both \verb|LSTM| and \verb|CNN| models. Second, we discover the \textbf{hidden long-term dependencies} and demonstrate experimentally that aperiodic time-series data not only rely on external factors but also exhibit dependencies with past period. Third, we introduce an \textbf{efficient model parameter aggregation approach} by employing ML algorithm in place of traditional ensemble models for combining parameters from multiple machine learning models. Finally, we validate our method through experiments on exchange rate datasets and show its effectiveness in comparison to a SOTA model, \verb|NHITS|.

@@

Như đã đề cập, \verb|LSTM| là mạng neural rất phổ biến trong việc handle các bài toán liên quan đến dự đoán trên dữ liệu time-series. \verb|LSTM| được sử dụng phổ biến như vậy bởi vì nó xử lý tốt vấn đề vanishing gradient (dễ dàng bắt gặp khi sử dụng Recurrent neural network) và có thể khai thác hiệu quả các mối quan hệ phi tuyến trong dữ liệu. Thật vậy, bằng cách duy trì cell-state trong mỗi iteration, \verb|LSTM| có thể khắc phục vấn đề vanishing gradient, từ đó bảo toàn khả năng capture các phụ thuộc dài hạn \cite{cheng2018applied}. Ngoài ra, \verb|LSTM| thực hiện rút trích đặc trưng với các hàm kích hoạt phi tuyến, giúp các tham số mô hình có thể capture tính phi tuyến của dữ liệu \cite{he2016exploiting}. Hai yếu tố nêu trên khiến cho \verb|LSTM| trở thành lựa chọn đầu tiên được nghĩ đến khi giải quyết các bài toán trên dữ liệu time-series.

% As aforementioned, \verb|LSTM| is a well-known neural network for handling prediction problems on time-series data. \verb|LSTM| is commonly used because it efficiently handles the problem of vanishing gradients (easily encountered when using Recurrent neural networks) and can effectively exploit nonlinear relationships in data. Indeed, by maintaining the cell-state in each iteration, \verb|LSTM| can overcome the vanishing gradient problem, thereby preserving the ability to capture long-term dependencies \cite{cheng2018applied}. In addition, \verb|LSTM| performs feature extraction with nonlinear activation functions, which helps the model parameters capture the nonlinearity of the data \cite{he2016exploiting}. These factors make \verb|LSTM| to be the first choice to think of when solving problems on time-series data.

@@

Mạng \verb|CNN| được sử dụng rất nhiều trong các tác vụ xử lý hình ảnh \citep{naranjo2020review, sharma2018analysis} bởi khả năng tổng hợp các quan hệ cục bộ. Không chỉ vậy, \verb|CNN| còn được dùng rất nhiều trong các tác vụ xử lý dữ liệu time-series như speech recognition \cite{dua2022developing}, natural language processing \cite{varshitha2023natural}. Điều đó chứng tỏ được khả năng của \verb|CNN| trong việc khám phá mối quan hệ thời gian cục bộ giữa các mẫu dữ liệu. Mặc dù vậy, \verb|CNN| lại rất ít được dùng trong các tác vụ dự đoán stock price hay foreign exchange. Trong nghiên cứu này, chúng tôi tận dụng khả năng rút trích đặc trưng cục bộ tuyệt vời của \verb|CNN| để tích hợp thêm thông tin ẩn vào quá trình huấn luyện của mô hình.

% \verb|CNN| is widely used in image processing tasks \citep{naranjo2020review, sharma2018analysis} because of its ability to synthesize local relationships. Not only that, \verb|CNN| is also widely used in time-series data processing tasks such as speech recognition \cite{dua2022developing}, natural language processing \cite{varshitha2023natural}. This proves the ability of \verb|CNN| in discovering local temporal relationships between data samples. However, \verb|CNN| is rarely used in aperiodic time-series data prediction tasks. In this study, we take advantage of \verb|CNN|'s excellent local feature extraction ability to incorporate more hidden information into the model training process.

@@

Các thuật toán Meta-learning (ML) dựa trên tối ưu, điển hình là MAML \cite{finn2017model} được biết đến với khả năng huấn luyện một mô hình có tính tổng quát cao, thích ứng nhanh trên tập dữ liệu mới thông qua một lượng nhỏ dữ liệu và số bước huấn luyện \citep{hospedales2021meta, vettoruzzo2024advances}. Với khả năng này, ML được sử dụng rất nhiều trong các tác vụ đòi hỏi khả năng đáp ứng của mô hình trên dữ liệu (e.g. cá nhân hóa mô hình học \citep{chen2018federated, fallah2020personalized,nguyen2022meta}, domain adaptation trong online learning \citep{hu2023meta, khoee2024domain}).

% Optimization-based Meta-learning (ML) algorithms, typically MAML \cite{finn2017model} are known for their ability to train a highly general, adaptive model on new datasets with a limited amount of data and a small number of training steps \citep{hospedales2021meta, vettoruzzo2024advances}. With this ability, ML is widely used in tasks that require the model's ability to adapt to the data (e.g. personalization of learning models \citep{chen2018federated, fallah2020personalized,nguyen2022meta}, domain adaptation in online learning \citep{hu2023meta, khoee2024domain}).
